---
title: "PREP Amplitudes"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: false
    theme:
      bg: "#202123"
      fg: "#B8BCC2"
      primary: "#EA80FC"
      secondary: "#00DAC6"
      base_font:
        google: Prompt
      heading_font:
        google: Proza Libre
---

### Zachary Traylor's Master's Thesis
[Return to Main Page](main.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ggpubr)
library(car)
library(multcomp)
library(tidyverse)
library(rstatix)
library(rcompanion)
library(emmeans)
library(readr)
library(ggplot2)
library(afex)
```

------------------------------------------------------------------------

NOTE: At the request of Dr. He, the ERP values **are** baseline corrected for this analysis. The baseline period of -1000-0ms was chosen as this aligned with the baseline used in the permutation test (see [Replication of Source Study](replication.html)). While this is a rather long time window, the baseline was calculated as the average amplitude throughout this window, and did not differ significantly from a smaller baseline of -200-0ms, so the larger window was chosen for the sake of consistency.

```{r message=FALSE, warning=FALSE}
erp1 <- read_csv("Data/Perception_erp.csv", col_types = cols(...1 = col_skip(), 
                Sex = col_factor(levels = c("male", "female")), 
                Stimulus = col_factor(levels = c("1", "2", "3"))))
erp1$Condition <- "Perception"
erp2 <- read_csv("Data/EDA_erp.csv", col_types = cols(...1 = col_skip(), 
                Sex = col_factor(levels = c("male", "female")), 
                Stimulus = col_factor(levels = c("1", "2", "3"))))
erp2$Condition <- "EDA"
erp3 <- read_csv("Data/Motor_erp.csv", col_types = cols(...1 = col_skip(), 
                Sex = col_factor(levels = c("male", "female")), 
                Stimulus = col_factor(levels = c("1", "2", "3"))))
erp3$Condition <- "Motor"
erp4 <- read_csv("Data/Control_erp.csv", col_types = cols(...1 = col_skip(), 
                Sex = col_factor(levels = c("male", "female")), 
                Stimulus = col_factor(levels = c("1", "2", "3"))))
erp4$Condition <- "Control"
erp <- do.call("rbind", list(erp1, erp2, erp3, erp4))

erp <- erp[!duplicated(erp), ]  # remove duplicate rows
erp <- subset(erp, erp$ID!=18)  # exclude subject 18 (DNF motor)
erp <- na.omit(erp)             # remove any NAs

# baseline correction, if desired
new <- data.frame(ID=NA,
                  Sex=NA,
                  Stimulus=NA,
                  Component=NA,
                  Value=NA,
                  Condition=NA)
for (subject in unique(erp$ID)) {  # for each subject
  for (stimulus in unique(erp$Stimulus)) {  # for each stimulus level
    for (condition in unique(erp$Condition)) {  # for each condition
      selection <- subset(erp, (erp$ID==subject & erp$Stimulus==stimulus & erp$Condition==condition))
      baseline <- subset(selection, selection$Component=="Baseline_Amp")$Value  # grab baseline amplitude
      selection$Value <- selection$Value - baseline  # subtract baseline from all values
      # NOTE: this also subtracts baseline from itself AND the latency value!
      
      new <- rbind(new, selection)
    }
  }
}
new <- na.omit(new)  # remove extra NA row at the top
erp <- new  # replace ERP with baseline corrected values
```

## Comparison Between Sexes

Finally, we take the previously loaded ERPdata and perform our main analysis. In order to analyze the effect of Sex on PREP amplitudes and latencies, we perform a mixed-effects ANOVA with Stimulus Intensity and Condition introduced as within-subject factors and sex as a between-subject factor.

```{r message=FALSE, warning=FALSE}
model <- function(df, title, post=FALSE, inter=FALSE) {
  temp <- subset(df, df$Component==title)
  temp <- temp[!(temp$Value %in% boxplot(temp$Value)$out),]  # remove rows with outliers
  
  full.model <- aov_ez(id = "ID", 
                       dv = "Value", 
                       data = temp, 
                       between = "Sex", 
                       within = c("Stimulus", "Condition"),
                       fun_aggregate = mean)

  if (post) {  # if we are running post-hoc...
    post.hoc <- emmeans(full.model, ~ Sex)
    print(summary(as.glht(pairs(post.hoc)), test=adjusted("bonferroni")))
  }
  if (inter) {  # if we have an interaction
    post.hoc <- emmeans(full.model, ~ Sex|Stimulus)
    print(summary(as.glht(pairs(post.hoc)), test=adjusted("bonferroni")))
  }
  
  # assumption testing
  print("Sample size:")
  print(length(temp$Value))
  print(table(temp$Sex))
  
  # 1.  Normality
  print("NORMALITY: Visually inspect for non-normal data")
  R = residuals(full.model)
  qqPlot(R)
  hist(R)
  
  # 2.  Homogeneity of Variance (Levene's test)
  print("HOMOGENEITY OF VARIANCE: For Levene's test, p of < 0.05 is a violation, but if there is little to no group imbalance (ex: between male/female) this is acceptable")
  print(leveneTest(Value ~ Sex * Stimulus * Condition, data = temp))
  print(table(temp$Sex))
  print(table(temp$Stimulus))
  print(table(temp$Condition))

  # 3. Sphericity
  print("Violations of sphericity have been Greenhouse-Geisser corrected")
  
  return(full.model)
}
```

$~$

### N1 variance explained by Sex

```{r message=FALSE, warning=FALSE}
m <- model(erp, "N1_Amp", inter=TRUE)
knitr::kable(nice(m))
afex_plot(m, x = "Stimulus", trace = "Sex", 
          panel = "Condition", error = "none", 
          mapping = c("color", "fill"), 
          data_geom = geom_boxplot, data_arg = list(width = 0.7), 
          point_arg = list(size = 1.5), line_arg = list(size = 1),
          dodge = 0.8) +
  ylab("N1 Amplitude (μV)") +
  scale_fill_manual(values = c("#2274A5", "#CE7DA5")) + 
  scale_color_manual(values = c("#2274A5", "#CE7DA5")) +
  scale_x_discrete(name = "Stimulus Level", labels = c("Low", "Med", "High"))

for (i in c("male", "female")) {
  for (j in c("1", "2", "3")){
    for (k in c("Perception", "EDA", "Motor", "Control")){
      m <- mean(subset(erp, erp$Component=="N1_Amp" & erp$Sex==i & erp$Stimulus==j & erp$Condition==k)$Value,
                na.rm = TRUE)
      s <- sd(subset(erp, erp$Component=="N1_Amp" & erp$Sex==i & erp$Stimulus==j & erp$Condition==k)$Value,
              na.rm = TRUE)
      print(sprintf("%s %s %s - %s (%s)", i, j, k, round(m, digits = 1), round(s, digits = 1)))
    }
  }
}

```

$~$

### N2 variance explained by Sex

```{r message=FALSE, warning=FALSE}
m <- model(erp, "N2_Amp")
knitr::kable(nice(m))
afex_plot(m, x = "Stimulus", trace = "Sex", 
          panel = "Condition", error = "none", 
          mapping = c("color", "fill"), 
          data_geom = geom_boxplot, data_arg = list(width = 0.7), 
          point_arg = list(size = 1.5), line_arg = list(size = 1),
          dodge = 0.8) +
  ylab("N2 Amplitude (μV)") +
  scale_fill_manual(values = c("#2274A5", "#CE7DA5")) + 
  scale_color_manual(values = c("#2274A5", "#CE7DA5")) +
  scale_x_discrete(name = "Stimulus Level", labels = c("Low", "Med", "High"))

for (i in c("male", "female")) {
  for (j in c("1", "2", "3")){
    for (k in c("Perception", "EDA", "Motor", "Control")){
      m <- mean(subset(erp, erp$Component=="N2_Amp" & erp$Sex==i & erp$Stimulus==j & erp$Condition==k)$Value,
                na.rm = TRUE)
      s <- sd(subset(erp, erp$Component=="N2_Amp" & erp$Sex==i & erp$Stimulus==j & erp$Condition==k)$Value,
              na.rm = TRUE)
      print(sprintf("%s %s %s - %s (%s)", i, j, k, round(m, digits = 1), round(s, digits = 1)))
    }
  }
}

```

$~$

### P2 variance explained by Sex

```{r message=FALSE, warning=FALSE}
m <- model(erp, "P2_Amp")
knitr::kable(nice(m))
afex_plot(m, x = "Stimulus", trace = "Sex", 
          panel = "Condition", error = "none", 
          mapping = c("color", "fill"), 
          data_geom = geom_boxplot, data_arg = list(width = 0.7), 
          point_arg = list(size = 1.5), line_arg = list(size = 1),
          dodge = 0.8) +
  ylab("P2 Amplitude (μV)") +
  scale_fill_manual(values = c("#2274A5", "#CE7DA5")) + 
  scale_color_manual(values = c("#2274A5", "#CE7DA5")) + 
  scale_x_discrete(name = "Stimulus Level", labels = c("Low", "Med", "High"))

for (i in c("male", "female")) {
  for (j in c("1", "2", "3")){
    for (k in c("Perception", "EDA", "Motor", "Control")){
      m <- mean(subset(erp, erp$Component=="P2_Amp" & erp$Sex==i & erp$Stimulus==j & erp$Condition==k)$Value,
                na.rm = TRUE)
      s <- sd(subset(erp, erp$Component=="P2_Amp" & erp$Sex==i & erp$Stimulus==j & erp$Condition==k)$Value,
              na.rm = TRUE)
      print(sprintf("%s %s %s - %s (%s)", i, j, k, round(m, digits = 1), round(s, digits = 1)))
    }
  }
}

```
